---
title: "Analysis of 2016 GDP Data: A Case Study"
author: "Tony Garza"
date: "June 20, 2016"
output: 
  html_document:
    keep_md: TRUE
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Introduction

The following R Markdown document is a case study in response to your inquiry about analyzing current GDP trends using data from The World Bank. (Original Data Sources: http://data.worldbank.org/data-catalog/GDP-ranking-table & http://data.worldbank.org/data-catalog/ed-stats). The data is current as of April (for GDP) and May (for Ed-Stats) of 2016.

This study will answer the following questions: <br/>
1. Match the data based on the country shortcode. How many of the IDs match? <br/>
2. Sort the data frame in ascending order by GDP rank (so United States is last). What is the 13th country in the resulting data frame? <br/>
3. What are the average GDP rankings for the "High income: OECD" and "High income: nonOECD" groups? <br/>
4. Plot the GDP for all of the countries. Use ggplot2 to color your plot by Income Group. <br/>
5. Cut the GDP ranking into 5 separate quantile groups. Make a table versus Income.Group. How many countries are Lower middle income but among the 38 nations with highest GDP? <br/>

Additionally, this R Markdown document will break down my processes for data gathering, data tidying, and data analysis used to draw conclusions to the questions stated above.<br/>

# Data Gathering Phase
```{r}
setwd("~/Dropbox/SMUDataScienceClass/1st class/Week6CaseStudy_Garza") # all important step of setting the working directory
site1 <- "https://d396qusza40orc.cloudfront.net/getdata%2Fdata%2FGDP.csv" # site1 houses the GDP data
download.file(site1, destfile = "./GDP.csv") # store it in GDP data into GDP.csv
site2 <- "https://d396qusza40orc.cloudfront.net/getdata%2Fdata%2FEDSTATS_Country.csv" # site2 houses the EdStats data
download.file(site2, destfile = "./edstats.csv") # store it in edstats.csv
```
Now we need to read the files into data frames in R using this code chunk:
```{r}
GDPData <- read.csv("GDP.csv", sep = ",", header = TRUE, stringsAsFactors = FALSE)  #read the GDP data into the GDPData data frame
EdStats <- read.csv("edstats.csv", sep = ",", header = TRUE, strip.white = TRUE,
    na.strings= c("NA", " ", "")) # Read the EdStats data into EdStats data frame; Note: I added some code after the face to replace spaces with NAs within EdStats; this is because much of the data here is non-numeric and I need to keep the strings intact; this seemed like the most efficient approach.
```
This may require you to install "gdata" package if not already installed (if necessary use the following commands):<br\>
install.packages("gdata")<br\>
library(gdata) <br\>

# Tidy the Data
Going to break this up into sections. First we'll work on tidying GDPData.

## Tidy GDPData
Now that we have the data read into data frames we need to tidy the data to be able to use it properly.
The following code chunk will tidy the data for "GDPData" data frame. First let's see what we have:
```{r}
head(GDPData)
str(GDPData)
```
The data is in a bit of a mess and no where close to how we need it. Also, all we need from this data frame is the country code ("country"), the rank ("rank"), and the GDP ("GDPinDollars") <br\>
Note: decided to use "country" as "countrycode" is an R function.
```{r}
country <- GDPData[1:103,1] #pull out the required country code data
#rank <- as.numeric(GDPData[1:103,2]) #pull out the required rank data and fix the fact that the data is not numeric
rank <- (GDPData[1:103,2])
GDPinDollars <- GDPData[1:103,5] #pull out the required GDP data
CleanGDPdata <- data.frame(country,rank,GDPinDollars) #create a new data frame with this data
CleanGDPdata$rank <- as.numeric(as.character(CleanGDPdata$rank)) # make rank numeric, need as character first because ran is a factor
CleanGDPdata$GDPinDollars <- as.numeric(gsub("[^[:digit:]]","", CleanGDPdata$GDPinDollars)) #make numeric to identify NAs
CountNAs <- sum(is.na(CleanGDPdata$GDPinDollars)) #need to figure out how many NAs we have in the data set
CountNAs #we have 4 NAs
```
There are 4 values in the GDPinDollars column that are NAs and I want to eliminate the rows associated with these NAs <br\>
```{r}
CleanGDPdata <- na.omit(CleanGDPdata) #elimnate rows where there are NAs
```
## Tidy EdStats Data
Now onto the EdStats data set. First let's see what we have:
```{r}
head(EdStats)
str(EdStats)
```
Let's pull out just the data we need from EdStats:
```{r}
country <- EdStats[1:234,1] #pull out the required country code data
Income.Group <- EdStats[1:234,3] #pull out the required rank data
CleanEdStats <- data.frame(country,Income.Group) #create a new data frame with this data
CountEdStatsNAs <- sum(is.na(CleanEdStats$Income.Group)) #need to figure out how many NAs we have in the data set
CountEdStatsNAs #we have 24 NAs within EdStats$Income.Group
```
We now know that we have 24 NAs in Income.Group. Now let's remove any rows that contain NAs:
```{r}
CleanEdStats <- na.omit(CleanEdStats) #elimnate rows where there are NAs
```
# Analyze the Data
## Merge the data based on country code. How many of the IDs matched?
```{r}
MergeGDPData <- merge(x=CleanGDPdata, y=CleanEdStats, by="country", all=TRUE) # Use merge function to bring the two data sets together by country
```
By merging the data we now have introduced NAs into the merged data set as, for example, there are countries in EdStats that are not in GDPData. Therefore, these countries will have an NA for rank and GDPinDollars. This is the clue needed to address the second part of the question: how many of the IDs matched?
```{r}
sum(!is.na(MergeGDPData$GDPinDollars)) #finding out how many rows in GDPinDollars are not NAs
```
This shows us that there are 99 IDs that matched.

## Sort the data frame in ascending order by GDP rank (so United States is last). What is the 13th country in the resulting data frame?
First let's go ahead and eliminate the rows with the new NAs.
```{r}
MergeGDPData <- na.omit(MergeGDPData) #elimnate rows where there are NAs
```
Actually sorting by the rank column in ascending order would put the United States first with a rank of "1". To make USA last it makes more sense to sort on GDPinDollars column:
```{r}
require(dplyr) #dplyr required for arrange function
MergeGDPData <- arrange(MergeGDPData,GDPinDollars) #this arranges the data in ascending order
MergeGDPData #display merged data
```
Now to determine what the 13th country is I use the following simple code:
```{r}
MergeGDPData$country[13] # tells me what the 13th value of the vector is
```
This shows that in this prescribed order, from least to greatest, Kenya is the 13th country.

## What are the average GDP rankings for the "High income: OECD" and "High income: nonOECD" groups? 
Using the subset function I separated out the "High income: OECD and nonOECD" portions of the data.
```{r}
subsetMergeDataOECD <- subset(MergeGDPData,Income.Group=="High income: OECD") # create subset using OECD
subsetMergeDataNonOECD <- subset(MergeGDPData,Income.Group=="High income: nonOECD") # create another subset using nonOECD
mean(subsetMergeDataOECD$rank) # calulate the mean of the rank data
```
This is the average GDP ranking for High Income: OECD subset.
```{r}
mean(subsetMergeDataNonOECD$rank) # calulate the mean of the rank data
```
This is the average GDP ranking for High Income: nonOECD subset.

## Plot the GDP for all of the countries. Use ggplot2 to color your plot by Income Group.
Using ggplot2 I plotted both a histogram and a qplot.
```{r}
require(ggplot2) # required package for ggplot and qplot
p <- ggplot(MergeGDPData,aes(GDPinDollars)) # ggplot
p + geom_histogram(fill="white",colour="black") #ggplot commands to display plot
```
As you can see in the histogram, the GDP data for all the countries is heavily skewed left to the side of more countries having very low GDPs. <br\>
```{r}

qplot(rank,GDPinDollars, data = MergeGDPData, color=Income.Group) #qplot of rank v GDPinDollars colored by Income.Group category
```
<br\>

The qplot shows an almost exponential curve with very few countries with high GDPs. It also shows some interesting data that some of the upper middle income countries are also some of the lowest ranking GDP producers. Conversersely, there are also a number of lower middle income countries that are high (or very high) GDP producers. 
```{r}
sum(MergeGDPData$GDPinDollars[1:94])
sum(MergeGDPData$GDPinDollars[95:99])
```
In fact this code proves that the top five countries in this data set together have a higher GDP than all the other countries combined.

## Cut the GDP ranking into 5 separate quantile groups. Make a table versus Income.Group. How many countries are Lower middle income but among the 38 nations with highest GDP?
Going to use quantile and cut functions along with dcast to create a table of ranking quantiles and Lower Middle Class data from Income.Group. 
```{r}
require(gtools)
rankcat <- quantile(MergeGDPData$rank, seq(0, 1, length = 6)) #This will give us five separate quantile groups
rankcat
require(reshape2) #need to load reshape2
MergeGDPData$quantile <- cut(MergeGDPData$rank, breaks = rankcat) 
dcast(MergeGDPData, quantile ~ Income.Group, length, value.var = "quantile")[, c("quantile", 
    "Lower middle income")] #create a table of quantile group vs Lower middle income from Income.Group
dcast(MergeGDPData, quantile ~ Income.Group, length, value.var = "quantile")[1, c("Lower middle income")] #calculate how many Lower Middle Income countries are also in top 38 countries in GDP.
```
There are a total of three Lower middle income countries that are also in the top 38 countries in GDP.

# Conclusion
In this case study we learned the severe difference between the production of the top highest producing countries in the world versus the production of the remainder of the world's countries. I gathered the most recent GDP data from the World Bank website along with data on household income. I then merged the two data sets to be able to gather additional insights and manipulate the data easier. I found that not only is the data severely skewed (as evident by the above histogram and qplot, it is almost exponential in the severity of its curve. The data was not ready to be used right away and this required me to condense and tidy the data. This included identifying and removing NAs, Making sure data was numerical (sometimes looked numerical but was really a character string), and condensing the data. I also removed extraneous rows, and renamed column names to match and assist in merging of data on a common variable. When these tidying tasks were complete I moved on to the analysis where I prepared the data for each analysis question asked. Here I used a combination of merge, ggplot, and quantile functions to come to assist in drawing conclusions about the data.